Supplemental Material - _metagenomeFeatures_: An R package for working with 16S rRNA reference databases and marker-gene survey feature data.

### 2018-11-08 inst/manuscript/supplemental_material.Rmd
revised text

---
title: "Supplemental Material - _metagenomeFeatures_: An R package for working with 16S rRNA reference databases and marker-gene survey feature data."
subtitle: "_Paenibacillus_ species resolution for 16S rRNA V12 and V4 regions."
author: "Nathan D. Olson and Nidhi Shah"
date: "`r Sys.Date()`"
output: bookdown::pdf_document2
---

@@

Here we demonstrate how _metagenomeFeatures_ and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions. 
Originally, _Paenibacillus_ was classified under the _Bacillus_ genus, a novel genus was formed based on the 16S rRNA gene similarity in the 1990s. 
_Paenibacillus_ spp. are facultative anaerobic bacteria present in a variety of environments including the soil, water, and can act as opportunistic pathogens in humans [@ouyang2008]. 
It has been shown that _Paenibacillus_ spp. will play an important role in sustainable agricultural industries [@grady2016]. 
Thus, an appropriate speciation of this genus is of an interest to the community.

@@

@@

```{r message = FALSE, echo = TRUE}
library(tidyverse); packageVersion("tidyverse")
library(ggpubr); packageVersion("ggpubr")
library(gplots); packageVersion("gplots")
library(DECIPHER); packageVersion("DECIPHER")
library(metagenomeFeatures); packageVersion("metagenomeFeatures")
library(greengenes13.5MgDb); packageVersion("greengenes13.5MgDb")
# library(silva128.1MgDb, lib.loc = "~/Documents/Fall2018/installed_packages/"); packageVersion("silva128.1MgDb")
library(silva128.1MgDb); packageVersion("silva128.1MgDb")
library(ribosomaldatabaseproject11.5MgDb); packageVersion("ribosomaldatabaseproject11.5MgDb")
```

# Database Comparison
- Three databases
  - greengenes
  - silva
  - Ribosomal database project

@@

# _Paenibacillus_ Sequence and Taxonomy Data
First subset the databases using the `mgDb_select()` function.
Then we will summarize the taxonomy data using functions from `tidyverse` package, specifically `dplyr`, `stringr` and `forcats` functions for manipulating `data.frames`, `strings`, and `factor` vectors respectively.

__TODO__ Clean-up do not want to load both taxa and seq into memory
```{r subsetDb, echo = TRUE}
paeni_16S_greengenes <- metagenomeFeatures::mgDb_select(gg13.5MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
paeni_16S_silva <- metagenomeFeatures::mgDb_select(slv128.1MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
paeni_16S_rdp <- metagenomeFeatures::mgDb_select(rdp11.5MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
```

```{r getCounts, echo = FALSE}
## Per genus count data - greengeness
tidy_gg_taxa <- paeni_16S_greengenes %>%
    ## cleaning up species names
    mutate(Species = if_else(Species == "s__", "Unassigned", Species),
           Species = str_replace(Species, "s__",""))

@@

## Per genus count data - SILVA
tidy_slv_taxa <- paeni_16S_silva %>% rowwise() %>% 
  mutate(new = str_split(Species, boundary("word")),Species = paste(new[1],new[2])) %>% 
  mutate(new = NULL) %>%
  mutate(Species = if_else(!str_detect(Species, "Paenibacillus"), "Unassigned", Species)) %>%
  mutate(Species = if_else(str_detect(Species, paste(c("Paenibacillus sp", "uncultured", "unidentified"), collapse = '|')), "Unassigned", Species)) %>%
  mutate(Species = str_remove(Species, "Paenibacillus "))

@@

## Per genus count data - RDP
tidy_rdp_taxa <- paeni_16S_rdp %>% rowwise() %>% 
  dplyr::rename(Species = species, Genus = genus) %>% 
  mutate(new = str_split(Species, boundary("word")), Species = paste(new[1],new[2])) %>% 
  mutate(new = NULL) %>%
  mutate(Species = if_else(!str_detect(Species, "Paenibacillus"), "Unassigned", Species)) %>%
  mutate(Species = if_else(str_detect(Species, paste(c("Paenibacillus sp", "uncultured", "unidentified"), collapse = '|')), "Unassigned", Species)) %>%
  mutate(Species = str_remove(Species, "Paenibacillus "))

@@

__TODO__ Revise text - all three databases
For the Greengenes 13.5 database, there are a total of `r total_otus_greengenes` sequences classified as `r nlevels(taxa_df_greengenes$Species)` species in the Genus _Paenibacillus_.
The number of sequences assigned to specific _Paenibacillus_ species range from 199 _Paenibacillus amylolyticus_ to 2 _Paenibacillus illinoisensis_ (Fig. \@ref(fig:speciesCount)).
Sequences only classified to the genus level, "Unassigned", is the most abundant group, `r no_species_assignment_greengenes`.

Similarly, we extracted _Paenibacillus_ sequences from Silva 128 database. Fig. \@ref(fig:speciesCountSilva) shows the number of species in the Genus _Paenibacillus_ in the Silva database.

For the RDP 11.5 database, all the sequences belonging to _Paenibacillus_ genus were either uncultured, unidentified, or unassigned.

```{r speciesCount, fig.cap = "Number of sequences in the Greengenes database assigned to species in the genus _Paenibacillus_ for the SILVA and Greengenes databases."}
combined_taxa_df <- bind_rows(list(Greengenes = taxa_df_greengenes, 
                                   SILVA = taxa_df_silva, 
                                   RDP = taxa_df_rdp), 
                              .id = "Database") %>% 
  group_by(Species) %>% 
  mutate(max_count = max(Count, na.rm = TRUE)) %>% 
  ungroup() %>% 
  top_n(60, max_count) %>% 
  mutate(Species = fct_reorder(Species, max_count))

@@

## Taxonomic resolution
Next, we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ Species by comparing within and between Species amplicon pairwise distance for the V12 and V4 regions.
To differentiate between Species the pairwise distances for within-Species amplicon sequences must be less than the between Species distances.
Additionally ,the difference in amplicon sequence pairwise distances between and within Species must be greater than the sequencing error rate to detect the difference.
For our taxonomic resolution analysis, we used pattern matching to extract the V12 and V4 regions of the 16S rRNA sequences.
We then generate a pairwise distance matrix for the two regions and compare the within and between Species pairwise distances.
Here, we work with sequences only in Greengenes 13.5 database for simplicity, the same analysis can be done with any other database.

For our _in-silico_ PCR we will use the following PCR primers:

@@

### Generating Distance Matrix for 16S Amplicons
Demonstrate process using gg and V12. 
Code hidden for other databases and V4, see Rmarkdown for full code, __PATH to Rmarkdown in repo__.

Extracting the V12 region from the database sequences, only sequences with containing both forward and reverse primers are included in the analysis.

@@

@@

## Excluding seqs with lenght 0
paeni_V12_seqs <- paeni_V12[[2]][width(paeni_V12[[2]]) != 0]
```

Generating a multiple sequence alignment using the `AlignSeqs` function in the `DECIPHER` package.

@@

Generating pairwise distance matrix using the `DistanceMatrix` function in the `DECIPHER` package for taxonomic resolution analysis and converting distance matrix to a data frame for analysis.

@@

### Multiple Database Comparison
```{r distDFfuns, echo = FALSE, message = FALSE}
get_amp <- function(paeni_16S, forward_primer, reverse_primer){

@@

#### Amplicon Sequence Lengths
The trimmed sequence length varies for forward and reverse primers resulting in varying amplicon sizes for both the V12 and V4 amplicons (Fig. \@ref(fig:trimInfo)).

```{r trimInfo, fig.cap = "Primer trimmed sequence, amplicon, length and start and end positions relative to full length sequences for the V12 and V4 regions."}
# list(V12 = paeni_V12[[1]], V4 = paeni_V4[[1]]) %>%
#     map_dfr(as.data.frame, .id = "amplicon") %>% 
#     filter(width != 0) %>% 
#     gather("key","value", -names, -amplicon) %>%
#     mutate(key = if_else(key == "width", "length", key),
#            key = factor(key, levels = c("length","start","end"))) %>%
#     mutate(position_type = if_else(key == "length",
#                                    "length", "position")) %>%
#     ggplot() +
#     geom_histogram(aes(x = value, fill = key), bins = 100) +
#     facet_grid(amplicon~position_type, scales = "free") +
#     theme_bw()
```

@@

Pairwise distance is significantly different for within and between Species comparisons indicating that the V12 and V4 regions are potentially suitable forclassifying members of the _Paenibacillus_ genus to the Species level (Fig. \@ref(fig:pairDist)).
Overall the V12 region had greater pairwise distances than V4 for both within and between Species.
It is important also to consider that the majority of sequences in the database were only classified to the genus level.
Species-level information for these sequences might yield results that are inconsistent with our analysis.
Additionally, our analysis does not identify the pairwise sequence distance required to classify a sequence as a novel _Paenibacillus_ Species.

@@

#### Species level comparison
While the overall pairwise distance is greater between Species than within Species for the _Paenibacillus_ genus, it is important to understand how the within and between Species pairwise distances compare for individual Species.
The heatmap below shows pairwise distance information for within and between different _Paenibacillus_ Species for the V12 and V4 regions (Fig. \@ref(fig:distHeatMap)).
Whether the sequences are assigned to more than one OTU depends on the pairwise sequence distance metric and linkage method employed by the clustering algorithm.
In general though for Species levels classification the maximum within Species distance should be less than the minimum between Species distance.
For example as the maximum within Species pairwise distance for _P. lentimorbus_ is 0.13 and the minimum between Species pairwise distance for _P. lentimorbus_ and _P. alvei_ is 0.08 (Fig. \@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two Species is not possible.

@@

## Conclusion
Here we demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, _greengenes13.5MgDb_, and other R packages, can be used to evaluate whether Species-level taxonomic classification is possible for a specific amplicon region.
The approach used here can easily be extended to use different 16S rRNA databases (starting with a different `MgDb`class object), taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).

@@

## Session Information
### System Information
```{r warning = FALSE}
s_info <- devtools::session_info(include_base = FALSE)
pander::pander(s_info$platform)
```

### Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>%
    knitr::kable(booktabs = TRUE)
```

>>>>

---
title: "Supplemental Material - _metagenomeFeatures_: An R package for working with 16S rRNA reference databases and marker-gene survey feature data."
author: "Nathan D. Olson and Nidhi Shah"
date: "`r Sys.Date()`"
output: bookdown::pdf_document2
---

@@

- Multiple 16S rRNA databases
- Can perform analysis on an individual database
- with metagenomeFeatures and MgDb annotation packages the same analysis can easily be performed on multiple databases, increasing the power of the analysis.
Here we demonstrate how _metagenomeFeatures_ and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions.
Originally, _Paenibacillus_ was classified under the _Bacillus_ genus, a novel genus was formed based on the 16S rRNA gene similarity in the 1990s.
_Paenibacillus_ spp. are facultative anaerobic bacteria present in a variety of environments including the soil, water, and can act as opportunistic pathogens in humans [@ouyang2008].
It has been shown that _Paenibacillus_ spp. will play an important role in sustainable agricultural industries [@grady2016].
Thus, an appropriate speciation of this genus is of an interest to the community.

@@

While the databases provide a consistent interface to the different 16S rRNA databases, the databases use different approaches to formating the taxonomic names.
As a result the majority of the code used to produce this document is tidying the data.
This code is not show in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures github repository, __ADD URL__

@@

```{r message = FALSE, echo = TRUE}
library(tidyverse); packageVersion("tidyverse")
library(ggpubr); packageVersion("ggpubr")
library(gplots); packageVersion("gplots")
library(DECIPHER); packageVersion("DECIPHER")
library(metagenomeFeatures); packageVersion("metagenomeFeatures")
library(greengenes13.5MgDb); packageVersion("greengenes13.5MgDb")
library(silva128.1MgDb); packageVersion("silva128.1MgDb")
library(ribosomaldatabaseproject11.5MgDb); packageVersion("ribosomaldatabaseproject11.5MgDb")
```

# 16S rRNA Database Comparison
# Database Comparison
- Three databases
  - greengenes
  - silva
  - Ribosomal database project

@@

# _Paenibacillus_ Taxonomic Resolution
For the taxonomic resolution analysis we will;
1. characterize the taxonomic composition of the _Paenibacillus_ genus in the three databases,
2. calculate pairwise distances between sequences for 30 species with the most sequences,
3. evaluate pairwise distances within and between species.

## Taxonomic Characterization
First obtain the taxonomic information for _Paenibacillus_ genus from the three databases using the `mgDb_select()` function.

```{r subsetDb, echo = TRUE}
paeni_16S_greengenes <- metagenomeFeatures::mgDb_select(gg13.5MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
paeni_16S_silva <- metagenomeFeatures::mgDb_select(slv128.1MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
paeni_16S_rdp <- metagenomeFeatures::mgDb_select(rdp11.5MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
```

```{r getCounts, echo = FALSE}
## Per genus count data - greengenes ###########################################
tidy_gg_taxa <- paeni_16S_greengenes %>%
    ## cleaning up species names
    mutate(Species = if_else(Species == "s__", "Unassigned", Species),
           Species = str_replace(Species, "s__",""))

@@

## Per genus count data - SILVA ################################################
tidy_slv_taxa <- paeni_16S_silva %>% rowwise() %>%
  mutate(new = str_split(Species, boundary("word")),Species = paste(new[1],new[2])) %>%
  mutate(new = NULL) %>%
  mutate(Species = if_else(!str_detect(Species, "Paenibacillus"), "Unassigned", Species)) %>%
  mutate(Species = if_else(str_detect(Species, paste(c("Paenibacillus sp", "uncultured", "unidentified"), collapse = '|')), "Unassigned", Species)) %>%
  mutate(Species = str_remove(Species, "Paenibacillus "))

@@

## Per genus count data - RDP ##################################################
tidy_rdp_taxa <- paeni_16S_rdp %>% rowwise() %>%
  dplyr::rename(Species = species, Genus = genus) %>%
  mutate(new = str_split(Species, boundary("word")), Species = paste(new[1],new[2])) %>%
  mutate(new = NULL) %>%
  mutate(Species = if_else(!str_detect(Species, "Paenibacillus"), "Unassigned", Species)) %>%
  mutate(Species = if_else(str_detect(Species, paste(c("Paenibacillus sp", "uncultured", "unidentified"), collapse = '|')), "Unassigned", Species)) %>%
  mutate(Species = str_remove(Species, "Paenibacillus "))

@@

- Relatively small overlap between the databases, multiple database analysis provide additional power.
- The databases vary in the number of sequences for each species with some databases not having any sequences for some species.
- Number of sequences and species for each database
- Greengenes benefit - cleaner species names
- SILVA benefit - additional sequences
- RDP benefit - additional sequences
- Issue with all databases, sequences only classified to the genus level, "Unassigned", is the most abundant group, Greeengenes - `r no_species_assignment_greengenes`, SILVA - `r no_species_assignment_silva`, and RDP - `r no_species_assignment_rdp` __FIG-REF__.

```{r speciesCount, fig.cap = "Number of sequences in the Greengenes database assigned to species in the genus _Paenibacillus_ for the SILVA and Greengenes databases."}
combined_taxa_df <- bind_rows(list(Greengenes = taxa_df_greengenes,
                                   SILVA = taxa_df_silva,
                                   RDP = taxa_df_rdp),
                              .id = "Database") %>%
  group_by(Species) %>%
  mutate(max_count = max(Count, na.rm = TRUE)) %>%
  ungroup() %>%
  top_n(30, max_count) %>%
  mutate(Species = fct_reorder(Species, max_count))

@@

## Taxonomic Resolution
Next, we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ Species by comparing within and between Species amplicon pairwise distance for the V12 and V4 regions.
To differentiate between Species the pairwise distances for within-Species amplicon sequences must be less than the between Species distances.
Additionally ,the difference in amplicon sequence pairwise distances between and within Species must be greater than the sequencing error rate to detect the difference.

### Generating Distance Matrix for 16S Amplicons
Here, we will demonstrate the process for extracting the target amplicon region and calculating pairwise distances using the Greengenes 13.5 database for simplicity.
See the source Rmarkdown file for code used calculate pairwise distances for other databases and the V4 region as well as tidying the data for downstream analysis.

The V12 and V4 regions were extracted from the database sequences using pattern matching.
Some sequences do not contain both forward and reverse primers as not all the  16S rRNA sequences are not full length.
Only sequences with both forward and reverse primers are included in the analysis.

The following PCR primers were used for our _in-silico_ PCR:

@@

@@

## Selecting Paenibacillus taxa and seq
paeni_16S_greengenes <- metagenomeFeatures::mgDb_select(gg13.5MgDb,
                          type = c("taxa", "seq"),
                          keys = "Paenibacillus",
                          keytype = "Genus")

@@

## Excluding seqs with length 0
paeni_V12_seqs <- paeni_V12[[2]][width(paeni_V12[[2]]) != 0]
```

For more accurate distance estimates a multiple sequence alignment is used to calculate pairwise distances.
We will use the `AlignSeqs` function in the `DECIPHER` package to generate the multiple sequence alignment.

@@

Next a pairwise distance matrix is generated using the `DistanceMatrix` function in the `DECIPHER` package for taxonomic resolution analysis and converting distance matrix to a data frame for analysis.

@@

### Multiple Database Comparison
Now that we have our pairwise distance matrix we can evaluate the taxonomic resolution of the genus _Paenibacillus_ by comparing the within and between species.
We will first compare the distribution of within and between species pairwise distances for the genus.
Then look at the pairwise distances within species __XYZ__ compared distances between __XYZ__ and other _Paenibacillu_ species.
To reduce the complexity of our analysis we are only going to look at the pairwise distances between sequences assigned to the __###__ species with the most sequences.
With more sequences we will have better estimates for within species pairwise distances but is likely to result in the exclusion of closely related species.

```{r distDFfuns, echo = FALSE, message = FALSE}
get_amp <- function(paeni_16S, forward_primer, reverse_primer){

@@

@@

Pairwise distance is significantly different for within and between Species comparisons indicating that the V12 and V4 regions are potentially suitable for classifying members of the _Paenibacillus_ genus to the Species level (Fig. \@ref(fig:pairDist)).
__TO ADD - Conclusions from multiple database analysis__.
Overall the V12 region had greater pairwise distances than V4 for both within and between Species.
It is important also to consider that we only included sequences with species assignments and the species with the most sequences in the database.
Species-level information for the unclassified sequences and inclusion of less well represented species might yield results that are inconsistent with our analysis.
Additionally, our analysis does not identify the pairwise sequence distance required to classify a sequence as a novel _Paenibacillus_ Species.

@@

#### Species level comparison
Add text for new figure

<!--
While the overall pairwise distance is greater between Species than within Species for the _Paenibacillus_ genus, it is important to understand how the within and between Species pairwise distances compare for individual Species.
The heatmap below shows pairwise distance information for within and between different _Paenibacillus_ Species for the V12 and V4 regions (Fig. \@ref(fig:distHeatMap)).
Whether the sequences are assigned to more than one OTU depends on the pairwise sequence distance metric and linkage method employed by the clustering algorithm.
In general though for Species levels classification the maximum within Species distance should be less than the minimum between Species distance.
For example as the maximum within Species pairwise distance for _P. lentimorbus_ is 0.13 and the minimum between Species pairwise distance for _P. lentimorbus_ and _P. alvei_ is 0.08 (Fig. \@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two Species is not possible.
-->

@@

# Conclusion
__TO ADD - Statement about database overlap and using RNAcentralIDs__
Here we demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, _greengenes13.5MgDb_, _silva128.1MgDb_, _ribosomaldatabaseproject11.5MgDb_, and other R packages, can be used to evaluate whether Species-level taxonomic classification is possible for a specific amplicon region.
The approach used here can easily be extended to use evaluate taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).

@@

# Session Information
## System Information
```{r warning = FALSE}
s_info <- devtools::session_info(include_base = FALSE)
pander::pander(s_info$platform)
```

## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>%
    knitr::kable(booktabs = TRUE)
```
